---
title: "Aliens Among U.S.: Analyzing UFO Sightings"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r, echo = FALSE, message = FALSE}
library(tidyverse)
library(leaflet)
library(leaflet.extras)
library(factoextra)

```

```{r, echo = FALSE, message = FALSE}
ufo = read_csv("./data/ufo_clean.csv")
```
## Motivation

Our project explores the reporting of unidentified flying objects (UFOs) throughout the U.S. UFOs, by their very nature, leave unanswered questions, some of which we hope to explore through this project. With the wealth of information gathered about these phenomena and motivated by the congressional hearings earlier this year, we wanted to create a resource to understand the data on UFO encounters. Our project addresses three key components to provide this understanding: when the encounters happened, where the encounters happened, and what was encountered.

## Related Work

Inspired by[National UFO Reporting Center](https://nuforc.org/), researches such as [Faculty perceptions of unidentified aerial phenomena](https://www.nature.com/articles/s41599-023-01746-3), [The Scientific Investigation of Unidentified Aerial Phenomena (UAP) Using Multimodal Ground-Based Observatories](https://www.worldscientific.com/doi/10.1142/S2251171723400068), our project seeks to expand on existing research using advanced statistical methods and R programming.

## Initial Questions

We started with key questions like:

- What are the hotspots for UFO sightings?

- How have reports of UFO sightings changed over time?

- Are there common features in these reports that can be quantitatively analyzed?

As our project progressed, we began to explore more nuanced questions.

## Evolution of Questions:

- Do certain geographical areas have higher frequencies of certain types of UFO sightings?

- Has the nature of UFO sightings changed over time?  For example, a follow-up to temporal pattern analysis, looking at changes in characteristics (shapes) of sightings across different decades.

- Are there correlations between UFO sightings and other variables?


## Data

Our primary work was on data collected by the National UFO Reporting Center (NUFORC). Timothy Renner pulled and hosted the data version we used separately. Separately. The version of the data that we used was pulled and separately hosted by Timothy Renner. We restricted our analysis to UFO reports made within the U.S.A. and removed reports that contained inconsistent locations (some reported U.S. but showed longitudes and latitudes in South America, Europe, or Canada) or were missing fundamental values. This left us with 100,820 reports from an initial set of 141,261 reports.

The duration of the UFO encounter was reported in the text and had inconsistent formatting, as well as unexpected special characters, e.g., “~8 minutes”, “30 SECS”, or “3 hours or more”. We wrote a function to identify numeric time values within these strings and the appropriate time units (seconds, minutes, hours) and convert these reports to numbers of seconds. While this approach wasn’t perfect, it allowed us to perform subsequent analyses on most of the data.

We incorporated data from a few additional sources as well. To locate cities that were near the location of UFO reports, we utilized a U.S. city dataset from simple maps. We used decennial census data from the U.S. Census Bureau to control for differing populations across states. We also incorporated basic information about U.S. geography from R and the GeoJSON dataset.


### Source:

- timothy renner
- Simple US cities data
- census
- us states + abbrev data
- GeoJSON
- NUFORC

## Exploratory Analysis

Our exploratory analysis included the following:

1.	Initial Data Summarization: 


Descriptive statistics: Calculate basic statistics like the number of sightings, average duration, and distributions of sightings over time. 

```{r, echo = FALSE, message = FALSE}
summary(ufo)

str(ufo)
```

Examples include:

-	Number of sightings: There are 100,802 reported UFO sightings in the dataset.

- The average duration of sightings: The average duration of these sightings is approximately 451819 seconds. Notice that this high average might be influenced by the dataset's high duration values, suggesting the need for further investigation and analyses.

-	Distributions of sightings over time: The distribution of sightings by year starts from 1969 and shows a consistent number each year. For the first few years, the counts are as follows:

-	1969: 8760 sightings
-	1970: 8760 sightings
-	1971: 8760 sightings
-	1972: 8764 sightings
-	1973: 8760 sightings

2.	Geographical Analysis:

Visualizations: Create heat maps to show the geographical distributions of sightings.


```{r, echo = FALSE, message = FALSE}
# Create Heat Maps
ggplot(ufo, aes(x = city_longitude, y = city_latitude)) +
  geom_bin2d(bins = 100) +
  scale_fill_gradient(low = "blue", high = "red") +
  labs(title = "Heat Map of UFO Sightings", 
       x = "Longitude", 
       y = "Latitude")
```


```{r, echo = FALSE, message = FALSE}
# Create Interactive Map
leaflet(ufo) |>
  addTiles() |>
  addHeatmap(lng = ~city_longitude,
             lat = ~city_latitude,
             intensity = -1, 
             radius = 15,
             blur = 20)
```

3.	Temporal Analysis:

Time Series Analysis: Analyze the frequency of sightings over time (years, months, and hours of the day).



4.	Categorical Analysis:

Bar Charts and Pie Charts: Examine the distribution of sightings by categories like the object's shape, duration, and text.

```{r, echo = FALSE, message = FALSE}
# Bar Chart for Shape
ufo |>
  count(shape) |>
  ggplot(aes(x = shape, y = n)) +
  geom_bar(stat = "identity", 
           fill = "blue") +
  labs(title = "Distribution of UFO Sightings by Shape",
       x = "Shape",
       y = "Number of Sightins") +
  theme_minimal()
```

```{r, echo = FALSE, message = FALSE}
# Pie Chart for Shape
shape_counts = ufo |>
  count(shape) 

pie(shape_counts$n,
    labels = shape_counts$shape,
    main = "Proportion of UFO Sightings by Shape", 
    col = rainbow(nrow(shape_counts)))
  
```
From the bar chart and pie chart for shape, we could notice one category stands out with a significantly higher number of sightings than the others, which is “light”. This indicates that rather than solid objects with distinct shapes, people are likelier to claim encounters with mysterious lights when they report UFOs. It’s also important to understand that “UFO” refers to any flying object the viewer can not identify. It doesn’t always indicate that it originated from another planet.


```{r, echo = FALSE, message = FALSE}
# Histogram for Duration
ufo |>
  ggplot(aes(x = duration_clean)) +
  geom_histogram(binwidth = 10,
                 fill = "blue",
                 color = "black") +
  labs(title = "HIstogram of UFO Sightings by Duration",
       x = "Duration",
       y = "Frequency") +
  theme_minimal()
```

Because the `duration `is very skewed and has a wide range in the histogram, we change it to categorize the durations into intervals and then plot a bar chart.


```{r, echo = FALSE, message = FALSE}
# Bar Chart for Duration
ufo = ufo |>
  mutate(duration_category = cut(duration_clean, 
                                 breaks = c(0, 30, 60, 120, 300, 600, Inf),
                                 labels = c("0-30", "30-60", "60-120", "120-300", "300-600", "600+"),
                                 right = FALSE))

ufo |>
  count(duration_category) |>
  ggplot(aes(x = duration_category, y = n)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(title = "Bar Chart of UFO Sightings by Duration Cattegories",
       x = "Duration Category",
       y = "Number of Sightings") +
  theme_minimal()
  
```


Short Duration Sightings (0–30 seconds): While casual viewers are likely to notice something strange for a brief amount of time, this group has a large frequency of sightings, which is typical.


The medium-duration sighting categories (60-120 and 120-300 seconds): exhibit a lower number of sightings compared to the shortest duration category. This could indicate that fewer UFOs are viewed for these prolonged periods of time, or that witnesses are less inclined to remain or report longer observations.

Long Duration Sightings (300-600 seconds): It's interesting to note that a notable rise has been observed in sightings that extend from five to ten minutes. There are several possible explanations for this, including the UFO's stable position, several observations recorded as a single event, or longer-lasting celestial phenomena misidentified as UFOs.

Very Long Duration Sightings (600+ seconds): This category also contains a high number of reports of sightings, some of which may be observations of events that are reported over an extended period of time or that linger all night.

NA:

5.	Correlation Analysis:

Statistical Tests: Conduct correlation analysis between variables like time of day, shape, duration, city_latitude, city_longitude, and sightings detail.



## Additional Analysis

We conducted formal statistical analyses, including:

###	Cluster Analysis: To identify hotspots.


```{r, echo = FALSE, message = FALSE}
# Elbow Method to determine k value
set.seed(123)

wss = sapply(1:10, function(k){
  kmeans(ufo[, c("city_latitude", "city_longitude")], k, nstart = 10)$tot.withinss
})

plot(1:10, wss, type = "b", pch = 19, frame = FALSE,
     xlab = "Number of clusters K",
     ylab = "Total within-clusters sum of squares")
```


```{r, echo = FALSE, message = FALSE}
# K-Means Clustering
set.seed(123)

ufo_clusters_2 = ufo |>
  select(city_latitude, city_longitude) |>
  na.omit() |>
  kmeans(centers = 2) 

ufo = ufo |>
  mutate(cluster_2 = ufo_clusters_2$cluster)
```


```{r, echo = FALSE, message = FALSE}
# Cluster Summary
cluster_summary_2 = ufo |>
  group_by(cluster_2) |>
  summarize(
    count = n(),
    avg_lagitude = mean(city_latitude, na.rm = TRUE),
    avg_longitude = mean(city_longitude, na.rm = TRUE)
  )
```


```{r, echo = FALSE, message = FALSE}
# Visualizing the Clusters
ufo |>
  ggplot(aes(x = city_longitude,
             y = city_latitude,
             color = as.factor(cluster_2))) +
  geom_point(alpha = 0.5) +
  labs(title = "UFO Sighting Hotspots",
       x = "Longitude",
       y = "Latitude") +
  theme_minimal()
```



Interpretation of 2 Clusters: Based on geographic location, the UFO encounters were divided into two separate groups by the k-means clustering method.

Based on the greater count and average latitude and longitude that revolve around critical urban locations, Cluster 1 may indicate sightings in more active or densely populated areas in the eastern US.

Cluster 2, with a smaller count and more dispersed average latitude and longitude, may represent sightings in less inhabited or rural locations in the western sections of the United States.

Eighty percent of Americans reside on the US East Coast. In contrast, just fifty percent of Americans dwell in the vast land west of the border, including California, home to sixty percent of the country's population.

These clusters might point to trends in UFO sighting reports, perhaps connected to population density, whereby urban areas report more sightings because there are more potential witnesses there; on the other hand, these could be areas with higher flight traffic, which means there are more opportunities to witness unidentified objects.



```{r, echo = FALSE, message = FALSE}
# 10 clusters
ufo_clusters_10 = ufo |>
  select(city_latitude, city_longitude) |>
  na.omit() |>
  kmeans(centers = 10) 

ufo = ufo |>
  mutate(cluster_10 = ufo_clusters_10$cluster)
```


```{r, echo = FALSE, message = FALSE}
# Cluster Summary
cluster_summary_10 = ufo |>
  group_by(cluster_10) |>
  summarize(
    count = n(),
    avg_lagitude = mean(city_latitude, na.rm = TRUE),
    avg_longitude = mean(city_longitude, na.rm = TRUE)
  )
```


```{r, echo = FALSE, message = FALSE}
# Visualizing the Clusters
ufo |>
  ggplot(aes(x = city_longitude,
             y = city_latitude,
             color = as.factor(cluster_10))) +
  geom_point(alpha = 0.5) +
  labs(title = "UFO Sighting Hotspots",
       x = "Longitude",
       y = "Latitude") +
  theme_minimal()
```


Interpretation of 10 Clusters: The sightings are grouped into more focused categories when there are ten clusters, which may indicate more subtle trends or hotspots.

The number of sightings may be impacted by some clusters associated with well-known urban regions and associated with increased human densities and technological presence.

Some clusters may be associated with isolated or rural areas, indicating a distinct type of UFO activity or reporting patterns less impacted by human activity.

Some sightings may be connected to military or experimental aircraft because specific clusters are close to airports, military sites, or research centers.
Some clusters are located close to borders and along coastlines. This may indicate a trend wherein more sightings are reported over sensitive areas or vast bodies of water.




###	Time Series Analysis: To understand trends over time.

###	Each analysis was accompanied by appropriate statistical tests to validate our findings, such as Chi-square test.


## Discussion

Findings:

The goal of our initial question was to locate UFO sighting hotspots. By using k-means clustering analysis, we were able to determine that sightings are more common in particularly isolated locales as well as heavily populated urban areas. This implies that while the probability of sighting reports rises with the number of possible witnesses, specific topographical characteristics or other unidentified factors can influence the locations of UFO sightings.

Regarding the typical features of UFO reports, our research showed various observed forms and actions. But reports of "lights," "circles," and "discs" were standard, and they fit with the traditional UFO imagery that has been widely accepted in the media and popular culture. This lends credence to the idea that reporting trends could be influenced by how the public views UFOs.

When exploring how reports of UFO sightings have changed over time, we conducted a temporal analysis that showed a noticeable increase in reported sightings during the late 20th century, which could be correlated with the rise in science fiction media during that time. The analysis also indicated a surge in reports following major cultural events, such as notable space missions and popular UFO-related movies.

According to our analysis, the majority of states saw a surge in UFO encounters around 2014. There are multiple possible explanations for this peak. More sightings and reports may have been caused by increasing public interest from significant celestial occurrences, increased availability of digital recording equipment, and the widespread use of drones. Social media's ability to spread information quickly contributed to this increase as well since sites like Facebook and Twitter have grown widely used and made it possible for sightings to be reported promptly.

What's interesting about 2018 is how few sightings there were. It also corresponds with when drone technology became much more accessible to the general public. This may have contributed to initial bewilderment but ultimately improved public awareness of these objects and reduced the number of UFO reports. Furthermore, modifications to the reporting and data collection regulations might have affected the quantity of reported sightings. One possible explanation for this reduction could have been a modification to the reporting guidelines of the National UFO Reporting Center.

The upsurge of sightings in 2019 and 2020 may be attributed to various environmental and socio-cultural variables. Global lockdowns resulted from the COVID-19 pandemic, which started at the end of 2019 and continued throughout 2020. People may have spent extra time at home due to the lockdowns, taking a closer look at the sky and witnessing things they might not have noticed. Furthermore, the public's newfound interest in aerial phenomena may have been sparked by the 2020 publication of the Pentagon's UFO movies, which in turn may have increased the number of sightings that have been reported.

Most sightings were short—often lasting less than two minutes—according to the recorded and categorized sighting durations, which contribute to the quantitative aspects of the reports. The ordinarily subjective character of UFO sighting accounts is given a veneer of objectivity by this numeric component.

Limitations:

Data Quality and Completeness:

-	The statistics on UFO sightings can be improved by complete reporting. It's possible that witnesses won't report every sighting or that their descriptions need to be revised.

-	The analysis may be skewed by errors in the reported data, such as misspelled dates, times, places, or durations.

Subjectivity in Sightings:

-	Because UFO sightings are dependent on firsthand reports, they are subjective. What one individual perceives as another may recognize a UFO as an everyday object or phenomenon.

-	The data does not consider the validity of the witnesses or the circumstances surrounding the sightings.

Reporting Bias:

-	Reporting bias could exist if sightings are more common in regions with greater interest in UFOs or where individuals are more inclined to report such incidents.

- The number of reports can increase or decrease depending on public interest and media coverage, and these fluctuations may not always correspond with real-world UFO activity.

Geospatial Analysis Limitations:

-	Geographic sightings may not conform to the spherical and uniformly sized cluster assumptions made by the k-means clustering algorithm.

- Mountain ranges, bodies of water, and urban vs. rural areas are geographic features that should be considered when using latitude and longitude coordinates.

Temporal Factors:

-	The temporal distribution of sightings may need to be considered in the analysis. For instance, if the data covers several decades, sighting frequencies may vary over time due to population growth or advancements in drones and other aircraft.

External Influences:

-	The data analysis does not consider external elements that can affect the chance of witnessing odd phenomena, such as weather, visibility, or celestial events.

Selection of Clusters:

-	The number of clusters in k-means is a somewhat arbitrary decision that can significantly impact the outcomes. The elbow technique, for example, may yield a partial answer for determining the ideal number of clusters.

Lack of Contextural Data:

-	It is difficult to make firm judgments regarding the type or cause of the sightings without more contextual information (such as the location of nearby airports or military installations).

Technological Changes:

-	Technological advancements may impact the quantity and nature of reported UFO sightings. The quantity and kind of UFO sightings can be influenced by technological advancements over time. For instance, the number of unexplained objects in the sky has increased due to the widespread use of drones.

Final Thoughts:

While the analysis provided valuable insights into the patterns of UFO sightings, it is essential to approach the findings cautiously. The subjectivity inherent in UFO reporting, coupled with the limitations of the clustering methods, means that our conclusions should be seen as preliminary. Further research, using more sophisticated geospatial analysis techniques and incorporating additional contextual data, is required to deepen our understanding of UFO sightings.


---


[Back to Home](index.html)