---
title: "Trends in Space"
output: html_document
---
### Load packages
```{r load packages and default options, message = FALSE}
# Load packages
library(tidyverse)
library(knitr)
library(leaflet)
library(usmap)

# Set default figure options
knitr::opts_chunk$set(
  fig.width = 6,
  out.width = "90%"
)

theme_set(theme_bw() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```


```{r data import, include = FALSE}
df_ufo = read_csv("data/ufo_clean.csv")
```

Where are UFO sightings reported? We plotted 1% of the reports and where they occurred throughout the U.S.

```{r}
set.seed(1)

df_ufo |> 
  sample_frac(0.01) |> 
  leaflet() |> 
  addProviderTiles(providers$CartoDB.Positron) |> 
  addCircleMarkers(~city_longitude,~city_latitude, radius = 1) 
```

Most common closest city

#### Table 1. Cities with most sightings nearby
```{r table with close cities}
df_ufo |> 
  group_by(closest_city) |> 
  summarize(n_obs = n()) |> 
  arrange(desc(n_obs)) |> 
  head(10) |> 
  kable()
```
The two cities with the most UFO sightings nearby are are Seattle, WA and Portland, OR. These are large cities but not the largest in the U.S. The West Coast seems to have more reports than other areas, though St. Louis is a bit of an outlier in this regard.

### Rural vs Urban Differences
Whether
```{r urban vs rural}
df_ufo |> 
  group_by(location, shape) |> 
  summarize(n_obs = n()) |> 
  arrange(desc(n_obs)) |> 
  pivot_wider(
    names_from = "location",
    values_from = "n_obs"
  ) |> 
  head(10) |> 
  kable()
```


### U.S. Heatmap

```{r us heatmap}
df_pop = 
  read_csv("data/us_census.csv") |>
  rename(state = abbrv) |> 
  select(state, census_2010)

df_ufo |> 
  group_by(state) |> 
  summarize(n_obs = n()) |> 
  left_join(df_pop, by = join_by(state)) |> 
  mutate(obs_per = n_obs/census_2010*100000) |> 
  plot_usmap(data = _, values = "obs_per", color = "#333333") +
  labs(
    title = "UFO Reports per 100,000 Population"
  ) + 
  scale_fill_continuous(name = "Reports per 100k") +
  theme(legend.position = "bottom")
```

