---
title: "Trends in Space"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

In this page we will explore where UFO sightings are reported in the United States. First we will visualize where throughout the U.S. these reports are being made, and then compare reports made in urban/suburban areas (i.e. close to major cities), and those made in rural areas (i.e. far from major cities). We separate these two groups using a cutoff criteria of being within 5 miles of a city with more than 200,000 residents.

```{r load packages and default options, message = FALSE}
## Default packages & settings
# Load packages
library(tidyverse)
library(knitr)
library(leaflet)
library(usmap)

# Set seed for reproducibility
set.seed(1)

# Set default figure options
knitr::opts_chunk$set(
  fig.width = 6,
  out.width = "90%"
)

theme_set(theme_bw() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

## Data import
# Import UFO data
df_ufo = read_csv("data/ufo_clean.csv")

# Add population data
df_pop = 
  read_csv("data/us_census.csv") |>
  rename(state = abbrv) |> 
  select(state, census_2010)
```

## Reports across the U.S.
### Raw Data
To begin, we simply visualize a random 2% of these reports to get a sense of where these sightings are throughout the country.

```{r 10 percent plot}
df_ufo |> 
  sample_frac(0.02) |> 
  leaflet() |> 
  addProviderTiles(providers$CartoDB.Positron) |> 
  addCircleMarkers(~city_longitude,~city_latitude, radius = 1) 
```


At first glance, these reports seem to cluster along the coasts and in other regions of high population density. So, we will adjust for population at the state level and re-visualize the results.

### Per-capita distribution
```{r us heatmap}
df_ufo |> 
  group_by(state) |> 
  summarize(n_obs = n()) |> 
  left_join(df_pop, by = join_by(state)) |> 
  mutate(obs_per = n_obs/census_2010*100000) |> 
  plot_usmap(data = _, values = "obs_per", color = "#333333") +
  labs(
    title = "UFO Reports per 100,000 Population"
  ) + 
  scale_fill_continuous(name = "Reports per 100k") +
  theme(legend.position = "bottom") 
```

This heatmap reveals three clusters with particularly high per-capita UFO reports. The Northwest, including Washington, Oregon, Idaho, Montana, and Alaska are the 5 states with the highest per-capita UFO reports ranging from 66 (AK) to 81 (WA) UFO reports per 100,000 residents. Next is northern New England, with New Hampshire, Vermont, and Maine taking positions 6-8 in the list. These states have between 61 (ME) and 66 (NH) UFO reports per 100,000. Rounding out the top 10, we have Arizona and New Mexico, with 61 and 59 reports per 100,000 residents, respectively.

#### Table 1. States with highest per-capita reports
```{r per capita table}
## Table form
df_ufo |> 
  group_by(state) |> 
  summarize(n_obs = n()) |> 
  left_join(df_pop, by = join_by(state)) |> 
  mutate(obs_per = n_obs/census_2010*100000) |> 
  arrange(desc(obs_per)) |> 
  head(10) |> 
  kable(
    col.names = c("State", "Number of UFO Sightings",
                  "Population (2010 census)", "UFO sightings per 100k"),
    digits = 1)
```


If we cluster the number of observations by nearest large city, we see a similar result. The two cities with the most UFO sightings nearby are are Seattle, WA and Portland, OR. These are large cities but not the largest in the U.S. The West Coast seems to have more reports than other areas, though St. Louis is a bit of an outlier in this regard.

#### Table 2. Cities with most sightings nearby
```{r table with close cities}
df_ufo |> 
  group_by(closest_city) |> 
  summarize(n_obs = n()) |> 
  arrange(desc(n_obs)) |> 
  head(10) |> 
  kable()
```

## Distance to cities
We also plotted the distribution of the `dist` variable, the distance to the nearest city of at least 200,000 residents. This histogram shows that the most reports occur close to large cities, but there is a long tail of reports that were made further away. There are also `r df_ufo |> filter(dist > 200) |> pull(dist) |> length()` reports that were made further than 200 miles from a large city, and `r df_ufo |> filter(dist > 400) |> pull(dist) |> length()` that were more than 400 miles away. 

```{r distance histogram}
df_ufo |> 
  filter(dist <= 200) |> 
  ggplot(aes(x = dist)) +
  geom_histogram(bins = 21, alpha = 0.7, col = I("black")) +
  labs(
    title = "Distribution of reports by distance",
    y = "Number of reports",
    x = "Distance to large city (miles)"
  ) 
```


<!-- put a histogram with of dist -->



## Rural vs Urban Differences
With this preliminary understanding of the geographic distribution of UFO reports, we wanted to explore if the distance from large cities is related to characteristics of the report. To do this, we separate reports into `urban/suburban` and `rural` categories, using a cutoff value of 5 miles  to a large city. This is not a perfect division, as this distinction should really be done by population density, but should be servicable to identify major differences, as will be seen below.

### UFO Shape
Our first question was whether the type of object reported differed between these group, as determined by the shape of the UFO.
```{r urban vs rural}
df_loc_shape = 
  df_ufo |> 
  group_by(location, shape) |> 
  summarize(n_obs = n()) |> 
  arrange(desc(n_obs)) |> 
  pivot_wider(
    names_from = "location",
    values_from = "n_obs"
  ) 

df_loc_shape |> 
  head(10) |> 
  kable(col.names = c("Shape", "Rural", "Urban/Suburban"))
```

To analyze whether descriptions of UFOs varied between these groups, we conducted a Chi-Squared test of homogeneity. This revealed that the distribution of shapes did differ significantly between the urban/suburban and rural categories.

```{r chisq shape loc}
df_loc_shape |> 
  select(-shape) |> 
  as.matrix() |> 
  chisq.test() 
```

### Encounter Duration

```{r}
df_ufo |> 
  drop_na(duration_clean) |> 
  filter(duration_clean != 0) |>
  ggplot(aes(x = duration_clean, fill = location)) +
  geom_histogram(alpha = 0.5, bins = 41, col = I("black")) +
  scale_x_continuous(
    trans = "log10",
    breaks = scales::trans_breaks("log10", function(x) 10^x),
    labels = scales::trans_format("log10", scales::math_format(10^.x))
                     ) +
  labs(
    title = "Distribution of encounter duration",
    y = "Number of reports",
    x = "Encounter Duration (seconds)",
    fill = "Location"
  )
```

```{r}
df_ufo |> 
  drop_na(duration_clean) |> 
  filter(duration_clean <= 7200) |> 
  t.test(duration_clean ~ location, data = _)
```

Even though the data are highly right-skewed, there are enough observations that it is still valid to compare the data using a t-test. The results of this test suggest that there is a significant difference between the average duration of the rural UFO sightings and the urban UFO sightings. We estimate that sightings in rural areas last for approximately one minute longer than sightings in urban areas (10.5 minutes vs. 9.5 minutes).

